---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyxl

[![Travis-CI Build Status](https://travis-ci.org/nacnudus/tidyxl.svg?branch=master)](https://travis-ci.org/nacnudus/tidyxl)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/nacnudus/tidyxl?branch=master&svg=true)](https://ci.appveyor.com/project/nacnudus/tidyxl)
[![codecov](https://codecov.io/github/nacnudus/tidyxl/coverage.svg?branch=master)](https://codecov.io/gh/nacnudus/tidyxl)

[tidyxl](https://github.com/nacnudus/tidyxl) imports non-tabular from Excel files into R.  It exposes cell content,
position, formatting and comments in a tidy structure for further manipulation,
especialy by the [unpivotr](https://github.com/nacnudus/unpivotr) package.  It
supports the xml-based file formats '.xlsx' and '.xlsm' via the embedded
[RapidXML](http://rapidxml.sourceforge.net) C++ library.  It does not support
the binary file formats '.xlsb' or '.xls'.

## Installation

```{r, echo = TRUE, eval = FALSE}
devtools::install_github("nacnudus/tidyxl")
```

## Examples

```{r, echo = TRUE}
library(readxl)
library(tidyxl)
```

The package includes a spreadsheet, 'examples.xlsx', which has a completely
unstructured and non-tabular layout. A popular package for importing
spreadsheets does the following:

```{r, echo = TRUE}
# This example is not run because readxl currently (2016-09-14) fails where
# any 'headers' are missing
# read_excel("./inst/extdata/examples.xlsx", "Sheet1")
```

[tidyxl](https://github.com/nacnudus/tidyxl) imports the same table into a
format suitable for non-tabular processing (see e.g. the
[unpivotr](https://github.com/nacnudus/unpivotr) package in 'Similar projects'
below).

```{r, echo = TRUE}
(pivoted <- contents("./inst/extdata/examples.xlsx", "Sheet1")[[1]])
```

Formatting is imported separately.

```{r, echo = TRUE}
# Only a small part of the formatting structure is shown here
formats("./inst/extdata/examples.xlsx")$local[1:3] %>% str
```

This enables subsetting of cells by their formats.

```{r, echo = TRUE}
(bold <- formats("./inst/extdata/examples.xlsx")$local$font$bold)
pivoted[pivoted$local_format_id %in% which(bold), ]
```

## Philosophy

Information in in many spreadsheets cannot be easily imported into R.  Why?

Most R packages that import spreadsheets have difficulty unless the layout of
the spreadsheet conforms to a strict definition of a 'table', e.g.:

* observations in rows
* variables in columns
* a single header row
* all information represented by characters, whether textual, logical, or
  numeric

These rules are designed to eliminate ambiguity in the interpretation of the
information.  But most spreadsheeting software relaxes these rules in a trade of
ambiguity for expression via other media:

* proximity (other than headers, i.e. other than being the first value at the
  top of a column)
* formatting (colours and borders)

Humans can usually resolve the ambiguities with contextual knowledge, but
computers are limited by their ignorance.  Programmers are hampered by:

* their language's expressiveness
* loss of information in transfer from spreadsheet to programming library

Information is lost when software discards it in order to force the data into
tabular form.  Sometimes date formatting is retained, but mostly formatting 
is lost, and position has to be inferred again.

[tidyxl](https://github.com/nacnudus/tidyxl) addresses the programmer's problems
by not discarding information.  It imports the content, position and formatting
of cells, leaving it up to the user to associate the different forms of
information, and to re-encode them in tabular form without loss.  Further
packages are planned to assist with that step.

## Similar projects

[tidyxl](https://github.com/nacnudus/tidyxl) was originally derived from
[readxl](https://github.com/hadley/readxl) and still contains some of the same
code, hence it inherits the GPL-3 licence.
[readxl](https://github.com/hadley/readxl) is intended for importing tabular
data with a single row of column headers, whereas
[tidyxl](https://github.com/nacnudus/tidyxl) is more general, and less clever.

The [rsheets](https://github.com/rsheets) project of several R packages is in
the early stages of importing spreadsheet information from Excel and Google
Sheets into R, manipulating it, and potentially parsing and processing formulas
and writing out to spreadsheet files.  In particular,
[jailbreaker](https://github.com/rsheets/jailbreakr) attempts to extract
non-tabular data from spreadsheets into tabular structures automatically via
some clever algorithms.

[tidyxl](https://github.com/nacnudus/tidyxl) differs from
[rsheets](https://github.com/rsheets) in scope
([tidyxl](https://github.com/nacnudus/tidyxl) will never import charts, for
example), and implementation ([tidyxl](https://github.com/nacnudus/tidyxl) is
implemented mainly in C++ and is quite fast, only a little slower than
[readxl](https://github.com/hadley/readxl)).
[unpivotr](https://github.com/nacnudus/unpivotr) is a package related to
[tidyxl](https://github.com/nacnudus/tidyxl) that provides tools for unpivoting
complex and non-tabular data layouts using I not AI (intelligence, not
artificial intelligence).  In this way it corresponds to
[jailbreaker](https://github.com/rsheets/jailbreakr), but with a different
philosophy.

## Roadmap

- [ ] Write tests
- [ ] Parse shared formulas and propagate to all associated cells.
- [ ] Propagate array formulas to all associated cells.
- [ ] Parse dates
- [ ] Detect cell types (date, boolean, string, number)
- [ ] Implement formatting import in C++ for speed.
